Measuring contrast processing in the visual system using SSVEP

### 2024-11-28 review.tex
Now on 'thresholds'

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\KOMAoption{captions}{tableheading}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

@@

The higher harmonics of the steady-state signal are generally thought to
reflect nonlinear processing in the visual system (Regan \& Regan,
1988). SSVEP signals can also be elicited by periodic changes of
stimulus properties other than achromatic and chromatic contrast, such
as motion, stereo depth, and facial identity or expression (see Norcia
\emph{et al.}, 2015, for an overview); however our focus here is on the
contrast response.

@@

Although its mathematical definition is straightforward, the
computations that underlie contrast processing in the brain have been
the subject of intense research for many decades. The neural code for
contrast, even in the earliest parts of visual cortex, is not simply a
linear transform of the contrast at the retina - instead, contrast
signals undergo a cascade of nonlinear processing stages that, broadly,
attempt to normalise the output relative to the spatiotemporal
environment. This normalization, achieved through a computation called
`contrast gain control' (Heeger, 1992; Foley, 1994; Carandini \& Heeger,
2011) maximises the sensitivity of the visual system by making optimal
use of neuronal bandwidth. As an example, a grating placed at the centre
of a low-contrast background typically appears more intense than the
same grating when superimposed on a high contrast background (see
Figure~\ref{fig-centresurround}; note that the code used to produce all
figures in this review is available in python and R at:
https://github.com/wadelab/contrastReviewPaper).

@@

@@

To illustrate the effect of contrast gain control (Heeger, 1992), we
next include a second component (a `mask') that contributes to the gain
pool of the first (the `target'). The mask stimulus will suppress the
response at the target frequency, reducing its amplitude. The
suppression is reciprocal - activity at the mask frequency is also
reduced by the presence of the target, in a contrast-dependent manner
(Busse \emph{et al.}, 2009). At a single pair of (matched) contrast
levels, we see a complex pattern of intermodulation terms in the
transduced waveforms (left panel of Figure~\ref{fig-twoinputs}) and in
the Fourier spectrum (middle panel of Figure~\ref{fig-twoinputs}). The
two components interact generating nonlinear intermodulation terms at
sums and differences of the input frequencies (F1, F2). The nature of
the interaction --- and therefore the pattern of intermodulation terms
--- is determined by the computations happening as the contrast signal
moves from retina to cortex (Regan \& Regan, 1988).

@@

\caption{\label{fig-twoinputs}Demonstration of how two inputs of
different frequencies are processed by a nonlinear transducer.
Additional harmonics are apparent in the Fourier spectrum (middle
panel), and the contrast response function is suppressed (right panel).}

@@

Even without considering a spatial component, the early visual system is
far more complex than the model here suggests. For example as well as
cells that code positive or negative contrast in a more or less
continuous manner, the retina also contains `transient' cells that code
temporal changes in contrast. These cells (Kuffler \emph{et al.}, 1957;
Alpern, 1971), and cells with similar properties in the LGN (Levitt
\emph{et al.}, 2001) and cortex (Hubel \& Wiesel, 1959; Movshon, 1975),
will introduce second harmonic components even when the stimulus itself
is modulated in an on-off fashion. Analogously, in the spatial domain,
so-called `simple cells' are sensitive to the polarity of a spatial
contrast modulation while `complex cells' respond to the presence of
patterned spatial contrast irrespective of its spatial position (Hubel
\& Wiesel, 1962). The SSVEP response to a contrast-reversing sine-wave
grating therefore contains information about nonlinear computations
performed across a range of retinal and cortical cell types.

The complexity of even a simple simulation of the frequency-domain
signal arising from non-linear interactions is intriguing. Presumably,
the signal measured from early visual cortex is the result of a cascade
of nonlinear retinal and cortical operations up to that point. It
therefore contains a `signature' or `fingerprint' of the computational
nature, order and parameters of those operations - including the shape
of the transducer functions and the computations involved in signal
combination. In principle, that information could be recovered from the
SSVEP signal - a possibility recognised in the early days of the
technique (Regan \& Regan, 1988). Although characterising the complete
set of computations along entire processing pathway is challenging,
careful parametric variation of the input stimuli does allow us to fit
models of early visual gain control using SSVEP data. For example, Tsai
\emph{et al.} (2012) demonstrated that a gain control model gave a good
account of the pattern of intermodulation responses produced by two
overlaid patterns flickering at different frequencies. This was achieved
by passing full stimulus waveforms through the transducer nonlinearity,
and calculating the Fourier spectrum of the model output. Our more
recent work on signal combination across eyes and space similarly
demonstrated close correspondence between the predictions of a
computational model and empirical data in humans (Baker \& Wade, 2017).
More detailed modelling of intracortical recordings (Groen \emph{et
al.}, 2022) has revealed details of the timecourse of gain control
effects, specifically that normalization is delayed slightly relative to
the initial visual response.

Similar changes to the contrast response function might also be obtained
using adaptation paradigms, in which the visual system is exposed to
high contrast stimuli for long durations. Psychophysically, adaptation
increases detection thresholds, but has little effect on contrast
discrimination performance (Ross \emph{et al.}, 1993), much like pattern
masks (Foley, 1994). Although SSVEP adaptation effects show strong
tuning for orientation (Campbell \& Maffei, 1970; Vergeer \emph{et al.},
2018) and spatial frequency (Mecacci \& Spinelli, 1976), there appear to
be no studies measuring the full SSVEP contrast response function before
and after an extended period of adaptation.

@@

Because SSVEP responses at detection threshold are very small,
estimating a threshold is achieved by measuring the contrast response
function at relatively high levels, and extrapolating back along the
function (either contrast vs response measured at a constant spatial
frequency or spatial frequency vs response at a constant contrast level)
to estimate its intercept with the x-axis (see
Figure~\ref{fig-sweepvep}). This contrast level was shown to correspond
approximately with psychophysically measured detection thresholds
(Norcia \emph{et al.}, 1986).

@@

\caption{\label{fig-sweepvep}Sweep VEP simulation showing how a contrast
detection threshold can be estimated from sweep VEP data measured at
higher contrasts. The solid line is the regression fit to the lowest
four data points, and the dashed line extrapolates the fit back to
determine the contrast value when y=0, which gives a threshold estimate
(red point).}

@@

A robust estimate of the threshold therefore requires the measurement of
the SSVEP amplitude at many different super-threshold contrast levels.
This was made faster by the development of the `sweep VEP' paradigm in
which the stimulus changed its contrast, spatial frequency or some other
property, throughout a trial (Tyler \emph{et al.}, 1979). To avoid
hysteresis effects, the sweep is sometimes conducted both up and down in
the same experiment (Tyler \emph{et al.}, 1979; Norcia \& Tyler,
1985\emph{a}; Norcia \emph{et al.}, 1990). The sweep VEP (really, a
sweep `SSVEP') technique is now commonly-used to obtain a rapid and
objective measurement of visual acuity. In particular, because of its
relative speed and simplicity, this technique has now become a standard
for conducting tests of visual acuity in very young subjects or where
behavioural tests are not appropriate (Ridder, 2004; Bach \emph{et al.},
2008; Hoffmann \emph{et al.}, 2017; Bach \& Farmer, 2020).

This approach has revealed much about the development of visual
abilities in infants (Harris \emph{et al.}, 1976; Atkinson \emph{et
al.}, 1979; Braddick \emph{et al.}, 1986). In general, SSVEP
measurements of infant vision have revealed that contrast sensitivity
for both achromatic and chromatic contrast as well as stereoscopic depth
perception develops earlier than had been supposed previously based on
behavioural readouts (Dobson \emph{et al.}, 1978; Norcia \& Tyler,
1985\emph{b}) with both chromatic and achromatic contrast detection
reaching near-adult levels by around six months. Spatial acuity as
measured by SSVEP reaches adult levels more slowly, but near-adult
levels are recorded around one year (Norcia \& Tyler, 1985\emph{b};
Hamer \emph{et al.}, 1989) compared to around six to seven years with
behavioural measures (Atkinson \& Braddick, 1983; Ellemberg \emph{et
al.}, 1999). At least some of this difference is likely due to the
relative objectivity and high SNR of the SSVEP technique compared to
other methods such as preferential looking, which require careful
measurement of the infant's gaze direction. However it should be noted
that other groups have reported electrophysiological correlates of
visual acuity that more closely match the behavioural measures (De
Vries-Khoe \& Spekreijse, 1982).

The SSVEP technique has also been used to study the development of the
contrast gain control mechanisms described in the previous section
(Candy \emph{et al.}, 2001; Pei \emph{et al.}, 2017). Although contrast
gain control is measurable in infants as young as six weeks old (Morrone
\& Burr, 1986; Skoczenski \& Norcia, 1998; Candy \emph{et al.}, 2001),
its development appears to be slower, with adult levels being reached at
approximately 11 years (Pei \emph{et al.}, 2017).

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

>>>>

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother

@@

\KOMAoption{captions}{tableheading}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

@@

The higher harmonics of the steady-state signal are generally thought to
reflect nonlinear processing in the visual system (Regan \& Regan,
1988). Because they can arise from different neuronal computations,
different populations and even different stages of the visual system,
different harmonics can exhibit different input-output functions. For
example, 1F and 2F responses can have different thresholds (Bobak
\emph{et al.}, 1984) and scalp topographies (Regan, 1973) while Kaester
et al (Kaestner \emph{et al.}, 2024) show both different CRF slopes and
thresholds for 1F and 2F components generated by dynamic noise.

SSVEP signals can also be elicited by periodic changes of stimulus
properties other than achromatic and chromatic contrast, such as motion,
stereo depth, and facial identity or expression (see Norcia \emph{et
al.}, 2015, for an overview); however our focus here is on the contrast
response.

@@

Although its mathematical definition is straightforward, the
computations that underlie contrast processing in the brain have been
the subject of intense research for many decades. The neural code for
contrast, even in the earliest parts of visual cortex, is not simply a
linear transform of the contrast at the retina - instead, contrast
signals undergo a cascade of nonlinear processing stages that, broadly,
attempt to normalise the output relative to the spatiotemporal
environment. This normalization, achieved through a computation called
`contrast gain control' (Heeger, 1992; Foley, 1994; Carandini \& Heeger,
2011) maximises the sensitivity of the visual system by making optimal
use of neuronal bandwidth. As an example, a grating placed at the centre
of a low-contrast background typically appears more intense than the
same grating when superimposed on a high contrast background (see
Figure~\ref{fig-centresurround}; note that the code used to produce all
figures in this review is available in python and R at:
https://github.com/wadelab/contrastReviewPaperVNS).

@@

Although the SSVEP does seem to reflect the activity of relatively
well-tuned neuronal populations , changing some aspect of the stimulus
may change the nature of those populations. For example, Campbell and
Maffei (Campbell \& Maffei, 1970) noted that densely-sampled
measurements of contrast responses can reveal the presence of two,
qualitatively different types of neurons that exhibit different
log-linear contrast functions (see also Souza \emph{et al.} (2007)).
Modeling such two-limbed contrast response functions using a single
sigmoidal function is therefore an approximation.

@@

To illustrate the effect of contrast gain control (Heeger, 1992), we
next include a second component (a `mask') that contributes to the gain
pool of the first (the `target'). The mask stimulus will suppress the
response at the target frequency, reducing its amplitude. The
suppression is reciprocal - activity at the mask frequency is also
reduced by the presence of the target, in a contrast-dependent manner
Busse \emph{et al.} (2009). At a single pair of (matched) contrast
levels, we see a complex pattern of intermodulation terms in the
transduced waveforms (left panel of Figure~\ref{fig-twoinputs}) and in
the Fourier spectrum (middle panel of Figure~\ref{fig-twoinputs}). The
two components interact generating nonlinear intermodulation terms at
sums and differences of the input frequencies (F1, F2). The nature of
the interaction --- and therefore the pattern of intermodulation terms
--- is determined by the computations happening as the contrast signal
moves from retina to cortex (Regan \& Regan, 1988).

@@

\caption{\label{fig-twoinputs}Demonstration of how two inputs of
different frequencies are processed by a nonlinear transducer.
Additional harmonics are apparent in the Fourier spectrum (middle
panel), and the contrast response function is suppressed (right panel).
The effects of using different nonlinearities can be explored in the
accompanying code.}

@@

Even without considering a spatial component, the early visual system is
far more complex than the model here suggests. For example as well as
cells that code positive or negative contrast in a more or less
continuous manner, the retina also contains `transient' cells that code
temporal changes in contrast. These cells (Kuffler \emph{et al.}, 1957;
Alpern, 1971), and cells with similar properties in the LGN (Levitt
\emph{et al.}, 2001) and cortex (Hubel \& Wiesel, 1959; Movshon, 1975),
will introduce second harmonic components even when the stimulus itself
is modulated in an on-off fashion (McKeefry \emph{et al.}, 1996).
Analogously, in the spatial domain, so-called `simple cells' are
sensitive to the polarity of a spatial contrast modulation while
`complex cells' respond to the presence of patterned spatial contrast
irrespective of its spatial position (Hubel \& Wiesel, 1962). Both the
amplitude and the phase of the SSVEP response to a single
contrast-reversing sine-wave grating therefore contains information
about nonlinear computations performed across a range of retinal and
cortical cell types.

The complexity of even a simple simulation of the frequency-domain
signal arising from non-linear interactions is intriguing. Presumably,
the signal measured from early visual cortex is the result of a cascade
of nonlinear retinal and cortical operations up to that point. It
therefore contains a `signature' or `fingerprint' of the computational
nature, order and parameters of those operations - including the shape
of the transducer functions and the computations involved in signal
combination. In principle, that information could be recovered from the
SSVEP signal - a possibility recognised in the early days of the
technique (Regan \& Regan, 1988). Although characterising the complete
set of computations along entire processing pathway is challenging,
careful parametric variation of the input stimuli does allow us to fit
models of early visual visual processing (e.g. (Zemon \& Gordon, 2006))
and, by incorporating mathematical models of gain control mechanisms
with either instantaneous, or temporally-integrated gain control, it is
possible to model gain control using SSVEP data. For example, (Tsai
\emph{et al.}, 2012) demonstrated that a relatively simple gain control
model gave a good account of the pattern of intermodulation responses
produced by two overlaid patterns flickering at different frequencies.
This was achieved by passing full stimulus waveforms through the
transducer nonlinearity, and calculating the Fourier spectrum of the
model output. Our more recent work on signal combination across eyes and
space similarly demonstrated close correspondence between the
predictions of a computational model and empirical data in humans (Baker
\& Wade, 2017). More detailed modelling of intracortical recordings
(Groen \emph{et al.}, 2022) has revealed details of the timecourse of
gain control effects, specifically that normalization is delayed
slightly relative to the initial visual response.

Similar changes to the contrast response function might also be obtained
using adaptation paradigms, in which the visual system is exposed to
high contrast stimuli for long durations. Psychophysically, adaptation
increases detection thresholds, but has little effect on contrast
discrimination performance (Ross \emph{et al.}, 1993), much like pattern
masks (Foley, 1994). Although SSVEP adaptation effects show strong
tuning for orientation (Campbell \& Maffei, 1970; Vergeer \emph{et al.},
2018) and spatial frequency (Mecacci \& Spinelli, 1976), there appear to
be very measurements of the full SSVEP contrast response function before
and after an extended period of adaptation. One exception is a study by
Bach et al (Bach \emph{et al.}, 1988) that examined the effects of
adaptation on SSVEP responses who found a paradoxical increase in the 1F
SSVEP amplitude following prolonged exposure to high-contrast stimuli
across much of the probe contrast range. They explain this as being due
to the presence of a nonlinear transducer function that saturates at
high probe contrast driving the response to higher harmonic frequencies.
Adaptation reduces the apparent contrast of the input, leading to less
distortion, a more sinusoidal output and a concentration of the
frequency-tagged response at lower harmonics. This explanation is
conceptually related to the `supersaturation' effect that we observe in
Figure 4.

@@

Because SSVEP responses at detection threshold are very small,
estimating a threshold is achieved by measuring the contrast response
function at relatively high levels (in the linear part of the
log-contrast response function prior to saturation), and extrapolating
back along the function (either contrast vs response measured at a
constant spatial frequency or spatial frequency vs response at a
constant contrast level) to estimate its intercept with the x-axis
(Campbell \& Maffei, 1970) (see Figure~\ref{fig-sweepvep}). This
contrast level has been shown to correspond approximately with
psychophysically measured detection thresholds Norcia \emph{et al.}
(1986). One factor affecting the accuracy of this technique is the
presence of noise in the EEG signal which will increase the apparent
amplitude of the SSVEP responses across the measurement range. The
effect of this noise is particulary apparent for recordings with low SNR
or where the slope of the CRF is shallow (and therefore where small
changes in the vertical offset of the fit result in large changes in the
X-axis intercept). Accurate estimates of thresholds therefore require
the experimenter to model this noise and compensate for its effects - as
detailed in (Norcia \emph{et al.}, 1989).

@@

\caption{\label{fig-sweepvep}Sweep VEP simulation showing how a contrast
detection threshold can be estimated from sweep VEP data measured at
higher contrasts. The solid line is the linear regression fit to the
lowest four data points prior to the satuating part of the curve, and
the dashed line extrapolates the fit back to determine the contrast
value when y=0, which gives a threshold estimate (red point).}

@@

A robust estimate of the threshold therefore requires the measurement of
the SSVEP amplitude at many different super-threshold contrast levels.
This was made faster by the development of the `sweep VEP' paradigm in
which the stimulus changed its contrast, spatial frequency or some other
property, throughout a trial Tyler \emph{et al.} (1979). To avoid
hysteresis or short-term adaptation effects, the sweep is sometimes
conducted both up and down in the same experiment (Tyler \emph{et al.},
1979; Norcia \& Tyler, 1985\emph{a}; Norcia \emph{et al.}, 1990). The
sweep VEP (really, a sweep `SSVEP') technique is now commonly-used to
obtain a rapid and objective measurement of visual acuity. In
particular, because of its relative speed and simplicity, this technique
has now become a standard for conducting tests of visual acuity in very
young subjects or where behavioural tests are not appropriate (Ridder,
2004; Bach \emph{et al.}, 2008; Hoffmann \emph{et al.}, 2017; Bach \&
Farmer, 2020).

The approach of estimating visual thresholds by extrapolating SSVEP
responses at higher stimulus levels has revealed much about the
development of visual abilities in infants (Harris \emph{et al.}, 1976;
Atkinson \emph{et al.}, 1979; Braddick \emph{et al.}, 1986). In general,
SSVEP measurements of infant vision have revealed that contrast
sensitivity for both achromatic and chromatic contrast as well as
stereoscopic depth perception develops earlier than had been supposed
previously based on behavioural readouts (Dobson \emph{et al.}, 1978;
Norcia \& Tyler, 1985\emph{b}) with both chromatic and achromatic
contrast detection reaching near-adult levels by around six months.
Spatial acuity as measured by SSVEP reaches adult levels more slowly,
but near-adult levels are recorded around one year (Norcia \& Tyler,
1985\emph{b}; Hamer \emph{et al.}, 1989) compared to around six to seven
years with behavioural measures (Atkinson \& Braddick, 1983; Ellemberg
\emph{et al.}, 1999). At least some of this difference is likely due to
the relative objectivity and high SNR of the SSVEP technique compared to
other methods such as preferential looking, which require careful
measurement of the infant's gaze direction. However it should be noted
that other groups have reported electrophysiological correlates of
visual acuity that more closely match the behavioural measures (De
Vries-Khoe \& Spekreijse, 1982).

The SSVEP technique has also been used to study the development of the
contrast gain control mechanisms described in the previous section.
Although response- (Morrone \& Burr, 1986) or contrast gain control
{[}Candy \emph{et al.} (2001); ; Skoczenski \& Norcia (1998){]} is
measurable in infants as young as six weeks old, its development appears
to be slower, with adult levels being reached at approximately 11 years
(Pei \emph{et al.}, 2017).

@@

\bibitem[\citeproctext]{ref-bachGreenleeBuhler1988}
Bach M, Greenlee M \& Buhler B (1988). Contrast adaptation can increase
visually evoked potential amplitude. \emph{Clinical Vision Sciences}
\textbf{3,} 185--194.

@@

\bibitem[\citeproctext]{ref-BobakIOVS1984}
Bobak P, Bodis-Wollner I, Harnois C \& Thornton J (1984). VEPs in humans
reveal high and low spatial contrast mechanisms. \emph{Investigative
Ophthalmology \& Visual Science} \textbf{25,} 980--983.

@@

\bibitem[\citeproctext]{ref-brownDevelopmentRivalryDichoptic1999}
Brown RJ, Candy TR \& Norcia AM (1999).
\href{https://www.ncbi.nlm.nih.gov/pubmed/10586959}{Development of
rivalry and dichoptic masking in human infants}. \emph{Investigative
Ophthalmology \& Visual Science} \textbf{40,} 3324--3333.

@@

\bibitem[\citeproctext]{ref-campbellVisualEvokedPotential1972}
Campbell FW \& Kulikowski JJ (1972).
\href{https://doi.org/10.1113/jphysiol.1972.sp009801}{The visual evoked
potential as a function of contrast of a grating pattern}. \emph{The
Journal of Physiology} \textbf{222,} 345--356.

@@

\bibitem[\citeproctext]{ref-KaestnerTVST2024}
Kaestner M, Chen YD, Clement C, Hodges A \& Norcia AM (2024).
\href{https://doi.org/10.1167/tvst.13.2.21}{Two disparity channels in
human visual cortex with different contrast and blur sensitivity}.
\emph{Translational Vision Science \& Technology} \textbf{13,} 21--21.

@@

\bibitem[\citeproctext]{ref-Nakayama1982}
Nakayama K \& Mackeben M (1982).
\href{https://doi.org/10.1016/0042-6989(82)90138-9}{Steady state visual
evoked potentials in the alert primate}. \emph{Vision Research}
\textbf{22,} 1261--1271.

@@

\bibitem[\citeproctext]{ref-Norcia1989}
Norcia AM, Tyler CW, Hamer RD \& Wesemann W (1989).
\href{https://doi.org/10.1016/0042-6989(89)90048-5}{Measurement of
spatial contrast sensitivity with the swept contrast {VEP}}.
\emph{Vision Research} \textbf{29,} 627--637.

@@

\bibitem[\citeproctext]{ref-reganRapidObjectiveRefraction1973}
REGAN D (1973). Rapid {Objective Refraction Using Evoked Brain
Potentials}. \emph{Investigative Ophthalmology \& Visual Science}
\textbf{12,} 669--679.

\bibitem[\citeproctext]{ref-reganEvokedPotentialsSpecific1973}
Regan D (1973).
\href{https://doi.org/10.1016/0042-6989(73)90237-X}{Evoked potentials
specific to spatial patterns of luminance and colour}. \emph{Vision
Research} \textbf{13,} 2381--2402.

@@

\bibitem[\citeproctext]{ref-Souza2007}
Souza GS, Gomes BD, Saito CA, Silva Filho M da \& Silveira LCL (2007).
Spatial luminance contrast sensitivity measured with transient {VEP}:
Comparison with psychophysics and evidence of multiple mechanisms.
\emph{Investigative ophthalmology \& visual science} \textbf{48,}
3396--3404.

@@

\bibitem[\citeproctext]{ref-Zemon2006Dynamic}
Zemon V \& Gordon J (2006).
\href{https://doi.org/10.1016/j.visres.2006.07.007}{Luminance-contrast
mechanisms in humans: Visual evoked potentials and a nonlinear model}.
\emph{Vision Research} \textbf{46,} 4163--4180.
